---
title: "Casual versus Annual Member Behavior Analysis Report"
author: "JohnPaul Medina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
source(here::here("config.R"))
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis explores usage patterns of Cyclistic’s bike-sharing service, aiming to uncover actionable insights that differentiate casual users from annual members. The goal is to support business strategies for increasing membership subscriptions.

### Objective:
- Identify differences in usage behavior between casual riders and annual members.
- Provide recommendations to convert casual riders into long-term subscribers.

## Business Problem

Cyclistic wants to increase its membership subscriptions. To support this goal, we need to analyze historical ride data and identify behavioral differences between casual riders and annual members.

### Business Task:
> **What key differences exist between casual riders and annual members, and how can we encourage casual riders to become annual members?**

## Data Source

The dataset used for this project comes from **[Divvy Bikes](https://divvybikes.com/system-data)**, a bike-share service operating in Chicago. The data is used in accordance with the [Divvy Bikes Data License Agreement](https://divvybikes.com/data-license-agreement).

The data includes:
- Ride ID, type of bike, and user category.
- Timestamps for the start and end of each ride.
- Starting and ending station locations, station ids, and names.

## Data Cleaning and Preparation

Before performing any analysis, it is essential to:
- Ensure data types are consistent across datasets.
- Convert date-time columns into the appropriate format.
- Identify and remove any inconsistencies, such as missing or duplicated entries.

### Initial Exploration of Datasets
1. Load the January 2024 dataset and analyze the column names and data type using str()
2. Create a function that reads each dataset and display all their column names.
3. Create a function that checks and displays each column's data type. 

#### Initial check of January 2024 dataset
Here we see that there are 13 columns and their data types. Notice that the 
started_at and ended_at columns are `characters` instead of dates in `POSIXct` 
which is the data type that will let us compute ride duration information. 
```{r Loading Jan 2024 and column atrributes check}
# Load the first dataset
jan_2024 <- read.csv(file.path(raw_data_dir,"202401-divvy-tripdata.csv"))

#Display the column names and their respective data types.
str(jan_2024)
```
#### Checking if every column in every dataset are consistent (same)
Next step is to check if all of the column names of every column in every 
dataset are the same.  
The result shows that all of the datasets have the same names through the use 
of the unique() function. 
```{r Column name check}
# Funciton that checks if all the column names are exactly the same
check_column_names <- function(directory){
  # Compile all the raw dataset into a list
  raw_files <- list.files(path = raw_data_dir, pattern = "*.csv",
                          full.names = TRUE)
  # Read the column names for each file
  column_names_list <- lapply(raw_files, 
                              function(file) colnames(read.csv(file, nrows=1)))
  
  # Check if all datasets have the same column names
  # unique function removes all duplicates except for the first one.
  unique_columns <- unique(column_names_list)
  
  if(length(unique_columns) == 1){ # If all are matching, only 1 left in the list
    print(" All datasets have consistent column names.")
  }else{
    print("Column name mismatch found!")
    print(unique_columns) # prints the unique columns
  }
}
# Call the function and provide the raw directory as the parameter
check_column_names(raw_data_dir)
```
#### Check if all the columns of each dataset have the same data type
To ensure consistency across all datasets used in this analysis, we verified that each column maintains the same data type across files.
Since missing values could affect type inference, they were explicitly set to NA to prevent inconsistencies.
```{r Column data type check}
# This function checks if all the data types in each column of every dataset
# is the same type. This will print the mismatched data types if found. 
check_data_types <- function(directory) {
  # List all CSV files
  files <- list.files(path = directory, pattern = "*.csv", full.names = TRUE)
  
  # Read and check column data types for each file
  data_types_list <- lapply(files, function(file) {
    df <- read.csv(file, na.strings = c("", "NA"))  # Treat empty strings as NA
    sapply(df, class)  # Get data types
  })
  
  # Check if all datasets have consistent data types
  unique_data_types <- unique(data_types_list)
  
  if (length(unique_data_types) == 1) {
    print("All datasets have consistent data types.")
  } else {
    print("Data type mismatch detected! Showing unique data types:")
    print(unique_data_types)
  }
}

# Run the function
check_data_types(raw_data_dir)

```
#### Combine the datasets into one dataset and repalce missing values with NA
Now that the datasets have been verified, they are now combined into a **single 
dataset** and saved as an rds and a csv in the `processed` folder.  
- **The RDS format (.rds)** is used for **faster access and memory efficiency** in R.  
- **The CSV format (.csv)** is also saved for **reference and external use**. 
```{r Combine dataset and replace missing values with NA}
# List all csv files from raw_data_dir
raw_files <- list.files(path = raw_data_dir, pattern = "*.csv", 
                        full.names = TRUE)

# Read all the files and replace empty data into NA
cyclistic_combined_raw <- lapply(raw_files, function(file){
  df <- read.csv(file, na.strings = c("","NA")) # convert missing values to NA
  return(df)
})

# Combine all datasets into one
combined_dataset <- bind_rows(cyclistic_combined_raw)

```
Verify the content of the combined data.
```{r Verify combined_data__raw content}
dim(combined_dataset)
head(combined_dataset)
tail(combined_dataset)

```

