---
title: "Cleaning Log - Cyclistic Customer Behavior Analysis"
author: "JohnPaul Medina"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Cleaning Log Introduction
This cleaning log is a demonstration and report of all the cleaning steps that 
were done to the dataset in the fulfillment of this analysis project.  
**NOTE: The scripts are loading the most up to date dataset, thus the 
verification and modification of the data are already reflected.**


## Config.R
This script was created to create a central location for loading libraries
and assigning variables for quick reference and access. 
```{r config.R, include = FALSE }
# Load required packages
library(here)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(knitr)
library(lubridate)

# Define base project directory, location of config.R considered base location
project_dir <- here::here()

# Define paths for raw and processed data
data_dir <- file.path(project_dir, "data")
raw_data_dir <- file.path(data_dir, "raw")
processed_data_dir <- file.path(data_dir, "processed")

# Define paths for scripts, reports, and outputs
scripts_dir <- file.path(project_dir, "scripts")
reports_dir <- file.path(project_dir, "reports")
output_dir <- file.path(project_dir, "output")

# Create directories if they don’t exist
dir.create(raw_data_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(processed_data_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(scripts_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(reports_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Define paths for processed dataset storage
combined_raw_data_file <- file.path(processed_data_dir, "cyclistic_combined_raw.rds")
cleaned_data_csv <- file.path(processed_data_dir, "cyclistic_cleaned.csv")
cleaned_data_rds <- file.path(processed_data_dir, "cyclistic_cleaned.rds")
cleaned_data_file <- file.path(processed_data_dir, "cyclistic_cleaned.csv")

# Define analysis parameters
time_bin_size <- 2  # Bin time-of-day into 2-hour intervals
timezone = "America/Chicago"

# Debug flag for printing directory paths
show_paths <- FALSE  # Set to TRUE to print paths

if (show_paths) {
  print(paste("Project Directory:", project_dir))
  print(paste("Raw Data Directory:", raw_data_dir))
  print(paste("Processed Data Directory:", processed_data_dir))
  print(paste("Scripts Directory:", scripts_dir))
  print(paste("Reports Directory:", reports_dir))
  print(paste("Output Directory:", output_dir))
}

```

## Initial Dataset Exploration
This section of the cleaning process was done to:  
1. Initial evaluation of the data, the columns, and column data types.  
2. After data structure familiarization, each of the dataset from January 2024 
to January 2025 were checked for column name consistency. 
3. After ensuring that every dataset contains the same column name, the next 
action done is to check if all the columns have consistent data type across all 
the datasets.
### First Look at the Dataset
```{r Explore Jan 2024 Dataset}
# Load the first dataset
jan_2024 <- read.csv(file.path(raw_data_dir,"202401-divvy-tripdata.csv"))

#Display the column names and their respective data types.
str(jan_2024)

```
### Column Name Check
This chunk comfirms that all the column names across all dataset ahve consistent 
names.
```{r Check Column Names Across All Dataset}
# Function that checks if all the column names are exactly the same
check_column_names <- function(directory){
  # Compile all the raw dataset into a list
  raw_files <- list.files(path = raw_data_dir, pattern = "*.csv",
                          full.names = TRUE)
  # Read the column names for each file
  column_names_list <- lapply(raw_files, 
                              function(file) colnames(read.csv(file, nrows=1)))
  
  # Check if all datasets have the same column names
  # unique function removes all duplicates except for the first one.
  unique_columns <- unique(column_names_list)
  
  if(length(unique_columns) == 1){ # If all are matching, only 1 left in the list
    print(" All datasets have consistent column names.")
  }else{
    print("Column name mismatch found!")
    print(unique_columns) # prints the unique columns
  }
}
# Call the function
check_column_names(raw_data_dir)
```
### Column Data Type Check
This chunk confirms that all the columns in every dataset have the same 
corresponding data types.
```{r Column Data Type Check}
# This function checks if all the data types in each column of every dataset
# is the same type. This will print the mismatched data types if found. 
check_data_types <- function(directory) {
  # List all CSV files
  files <- list.files(path = directory, pattern = "*.csv", full.names = TRUE)
  
  # Read and check column data types for each file
  data_types_list <- lapply(files, function(file) {
    df <- read.csv(file, na.strings = c("", "NA"))  # Treat empty strings as NA
    sapply(df, class)  # Get data types
  })
  
  # Check if all datasets have consistent data types
  unique_data_types <- unique(data_types_list)
  
  if (length(unique_data_types) == 1) {
    print("All datasets have consistent data types.")
  } else {
    print("Data type mismatch detected! Showing unique data types:")
    print(unique_data_types)
  }
}

# Run the function
check_data_types(raw_data_dir)
```
## Combining the Datasets and Setting Missing Data to NA
The chunk combines all the dataset into one dataset while also ensuring that 
every missing data is set to NA for consistency and data accuracy when the
analysis work begins.  
As seen in the output, the head and tail of the combined dataset shows data from 
January 2024 and January 2025.
```{r Combine Datasets and Set Missing Data to NA}
# List all csv files from raw_data_dir
raw_files <- list.files(path = raw_data_dir, pattern = "*.csv", 
                        full.names = TRUE)

# Read all the files and replace empty data into NA
cyclistic_combined_raw <- lapply(raw_files, function(file){
  df <- read.csv(file, na.strings = c("", "NA"))  # Convert "" and "NA" to NA
  return(df)
})

# Combine all datasets into one dataset
combined_dataset <- bind_rows(cyclistic_combined_raw)

# Script for saving data as RDS and CSV removed to avoid overwriting the 
# current data.

# Checking the head and tail to show the combined data
head(combined_dataset)
tail(combined_dataset)


```

## Convert Character Dates to POSIXct
The `started_at` and `ended_at` columns are character data, to facilitate 
computation for analysis, these columns are converted into POSIXct.
```{r Date Conversion}
# CLEANING_convert_character_dates_to_posixct
# THis script will convert the started_at and ended_at columns from character 
# to POSIXct to facilitate computation. 

# Load the raw data
raw_combined_data <- readRDS(combined_raw_data_file)

convert_dates <- function(df){
  df$started_at <- ymd_hms(df$started_at, tz = timezone)
  df$ended_at <- ymd_hms(df$ended_at, tz = timezone)
  return(df)
}

# Apply the function
processed_data <- convert_dates(raw_combined_data)

# Check if conversion was successful
str(processed_data$started_at)
str(processed_data$ended_at)

# Script for saving the cleaned dataset is removed to avoid overwriting the 
# project's current progress.
```
## Check for Missing Values for Each Column
The next step is to check for missing values in each column. This is an 
essential step in verifying data completeness and integrity.  
#### Evaluation
The chunk shows that about **20%** of the data contains missing 
start_station_id, start_station_name, end_station_name, and end_station_id. This 
amount of data is **not negligible**, fortunately, the absence of these 
information will not hinder our analysis of the `ride_duration`, so these data 
were not removed from the analysis pool. 
```{r Count Missing Data by Column}
# CLEANING_check_missing_values_per_column
# This script functions as a way to explore the data by tracking which 
# columns have missing data. 

# Load the most current version of the cleaned_data_rds to ensure accuracy
df <- readRDS(cleaned_data_rds)

# Function that counts the missing values per column
missing_values_summary <- function(df) {
  missing_counts <- sapply(df, function(x) sum(is.na(x)))
  
  # Convert it to a data frame for better readability
  missing_summary <- data.frame(Column = names(missing_counts),
                                Missing_Count = missing_counts)
  
  # Sort by highest missing values first
  missing_summary <- missing_summary[order(-missing_summary$Missing_Count),]
  
  return(missing_summary)
}

missing_summary <- missing_values_summary(df)

print(missing_summary)
```
## Computing and Adding the `ride_duration` Column
To gain an insight on casual and annual member behavior and usage, it is 
necessary to analyze their corresponding `ride_duration`. Since there are no 
missing `started_at` and `ended_at` data, all of the `ride_duration` rows 
will have a corresponding value.
```{r ride_duration Column}
# CLEANING_ride_duration_column
# This script will compute the ride duration based on started_at and 
# ended_at POSIXct values and add a column. 

# Load the most current version of the cleaned_data_rds to ensure accuracy
cleaned_data <- readRDS(cleaned_data_rds)

# Function that calculates the ride duration in terms of minutes
calculate_ride_duration <- function(df){
  df$ride_duration <- as.numeric(difftime(df$ended_at, df$started_at,
                                          units="mins"))
  
  # Flag invalid duration of negative or zero minutes
  df$invalid_duration <- df$ride_duration <=0
  
  # Count the number of flagged rows
  num_invalid <- sum(df$invalid_duration)
  
  cat("Found", num_invalid, "rides with invalid durations(<=0 minutes.\n")
  cat("invalid rides flagged but not removed. \n")
  
  return(df)
}

# Use the function
cleaned_dataset <- calculate_ride_duration(cleaned_data)

str(cleaned_data)

# Script for saving the new cleaned_dataset is removed to avoid overwriting
# current cleaned dataset progress. 

```
### `ride_duration` Summary
The initial look at the `ride_duration` summary shows that there are 
negative time duration and unusually high positive time duration. This shows 
that further investigation of the `time_duration` data is necessary to filter 
erroneous, invalid, and outlier data. 
```{r Time Duration Summary}
summary(raw_combined_data$ride_duration)
```





