---
title: "Cyclistic User Behavior Exploratory Data Analysis"
author: "JohnPaul Medina"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
library(here)
setwd(here::here())  # Forces knitting to use project root
source(here::here("simulation/simConfig.R"))
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis (EDA) Log Introduction

### Objective: 
This **Exploratory Data Analysis (EDA)** aims to identify key 
**behavioral differences** between casual riders and annual members, helping 
shape a marketing strategy to encourage casual riders to convert to membership.

### EDA Focus Areas
The analysis is structured around the following key focus points:

1. **User Distribution Analysis**

* Compare the **proportion of casual vs. member riders.**
* Identify the **scale of potential member conversions**

2. **Ride duration Patterns**

* Compare **summary statistics of ride duration** across user types.
* Determine whether casual **riders take longer trips** and how this insight can 
can be used for marketing.

3. **Time-Based Trends in Service Usage**

* **Hourly and weekly heatmaps*** to analyze user riding pattern at different 
times of the day and days of the week.
* **Seasonal trends** by analyzing monthly ride distributions to evaluate peak 
demand periods. 

4. **Bike Type Preference & Usage Patterns**

* Identify **bike type preference** between causal and member riders.
* Explore **ride duration patterns across different bike types.**

Each of these areas will provide insights to support **marketing strategies**, 
optimize **pricing model**, and improve **customer engagement initiatives.**

---

## Preliminary General Data Exploration**Distribution Analysis

First we will take a look at **rider type distribution** to understand the 
**size of our data** and how casual versus member riders are distributed.

```{r casual-versus-member-distribution-analysis}
if (!exists("cleaned_data")) {
  cleaned_data <- readRDS(cleaned_data_rds)
}

# Compute Rider Counts
member_count <- sum(cleaned_data$member_casual == "member")
casual_count <- sum(cleaned_data$member_casual == "casual")
total_accounts <- member_count + casual_count

# Create a summary table
rider_type_summary <- data.frame(
  Rider_Type = c("Casual Riders", "Member Riders", "Total Riders"),
  Count = c(casual_count, member_count, total_accounts),
  Percentage = c(
    (casual_count / total_accounts) * 100,
    (member_count / total_accounts) * 100,
    100
  )
)

# Print the result as a formatted table
knitr::kable(rider_type_summary,
             caption = "Rider Type Distribution",
             digits = 2,
             format = "pipe")
```
```{r member-vs-casual-distribution-visualization}
# Create a Data Frame for Plotting
account_counts <- data.frame(
  account_type = c("Casual", "Member"),
  count = c(casual_count, member_count)
)

# Plot Bar Chart of rider type distribution
p <- ggplot(account_counts, aes(x = account_type, 
                                y = count, fill = account_type)) +
  geom_bar(stat = "identity", width = 0.6) +  # Adjust width for aesthetics
  geom_text(aes(label = count), vjust = -0.5, size = 5, fontface = "bold") +  
  labs(
    title = "Distribution of Casual vs. Member Riders",
    x = "Account Type",
    y = "Number of Riders",
    fill = "Account Type"
  ) +
  scale_fill_manual(values = c("Casual" = "skyblue", "Member" = "darkblue")) +  
  theme_light() +  
  expand_limits(y = max(account_counts$count) * 1.1) +  
  theme(plot.margin = margin(10, 10, 10, 10))

# Display the Plot
print(p)
```
### Analysis and Key Insights

* There is an **uneven distribution** of casual(36.16%) versus member (63.84%)
riders. 
*This *high number of casual riders** presents an opportunity for 
**membership conversion marketing.**
* To ensure fair comparisons, **percentages or averages** should be used 
instead of raw counts to avoid skewed results due to the **imbalance between 
casual and member totals.**

---

Next we will conduct an analysis of ride duration between casual and member 
riders. 

```{r compare-summary-for-casual-vs-member}
# Study the summary of ride_duration of casual riders vs annual member riders
casual_summary <- 
  summary(cleaned_data$ride_duration[cleaned_data$member_casual=="casual"])
member_summary <- 
  summary(cleaned_data$ride_duration[cleaned_data$member_casual == "member"])


# Create a data frame with the summary statistics
summary_data <- data.frame(
  Statistic = rep(names(casual_summary),2),
  Value = c(as.numeric(casual_summary), as.numeric(member_summary)),
  Rider_Type = rep(c("Casual", "Member"), each = length(casual_summary))
)

# Convert the data frame to wide format for table display
ride_duration_summary <- tidyr::pivot_wider(
  summary_data,
  names_from = Rider_Type,
  values_from = Value
)

# Compute Percentage Difference only for Q1, Median, Mean, and Q3
ride_duration_summary$Percent_Difference <- ifelse(
  ride_duration_summary$Statistic %in% c("1st Qu.", "Median", "Mean", "3rd Qu."),
  round(((ride_duration_summary$Casual - ride_duration_summary$Member) / 
         ride_duration_summary$Member) * 100, 2),
  NA  # Leave Min and Max without percentage difference
)

# Display summary table
knitr::kable(ride_duration_summary,
             caption = "Ride Duration Summary by Rider Type",
             digits = 2,
             format = "pipe")
```

Bar graph of ride duration summary by rider type

```{r ride-duration-vs-rider-type-summary}
# Exclude "Max" and "Min" statistics from the data
summary_data <- summary_data %>% 
  filter(Statistic %in% c("1st Qu.", "Median", "Mean", "3rd Qu."))

# Ensure correct ordering of statistics
summary_data$Statistic <- factor(summary_data$Statistic,
                                 levels = c("1st Qu.", "Median", "Mean", "3rd Qu."))

# Create the dodge bar graph with labels
ggplot(summary_data, aes(x = Statistic, y = Value, fill = Rider_Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_text(aes(label = round(Value, 2)), 
            position = position_dodge(width = 0.8), 
            vjust = -0.3, size = 4) + # Adjust vjust for padding
  scale_fill_manual(values = c("Casual" = "skyblue", "Member" = "darkblue")) +
  labs(title = "Ride Duration Summary by Rider Type (Excluding Max and Min)",
       x = "Summary Statistic",
       y = "Ride Duration (Minutes)") +
  theme_light()
```
Box plot visualization

```{r ride-duration-vs-rider-type-box-plot}
# Box plot of the summaries of casual vs member ride durations
ggplot(cleaned_data, aes(x = member_casual, y = ride_duration, fill = member_casual)) +
  geom_boxplot() + 
  scale_y_log10() +  # Apply log scale to y-axis
  labs(
    title = "Ride Duration Comparison (Log Scale)",
    x = "Account Type",
    y = "Log Ride Duration (minutes)"
  ) +
  scale_fill_manual(values = c("casual" = "skyblue", "member" = "darkblue")) + 
  theme_light()
```
### Analysis and Key Inisghts

* Based on numerical and visual comparison of the summaries of casual and 
member ride duration, casual riders tend to use the bike longer than members.
* The box plot shows that casual riders tend to take longer rides than members.
* Member ride durations from Q1 to Q3 are contained within a **narrower range**, 
indicating a more **consistent** riding pattern.

---

Next we will analyze the riders' bike preference.

```{r ride-type-vs-rider-type-percentage}
# Compute the percentage of use of ride type based on rider type.
ride_type_percentage <- cleaned_data %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(count = n(), .groups="drop") %>% 
  group_by(member_casual) %>% 
  mutate(percent = (count / sum(count)) * 100) %>% 
  ungroup()

# Convert to a table for better readability
knitr::kable(
  ride_type_percentage,
  caption = "Bike Type Usage by Rider Type (Percentage Breakdown)",
  digits = 2,
  format = "pipe"
)
  
# Visualize the rideable_type vs member_casual as a bar graph
ggplot(ride_type_percentage, aes(x = rideable_type, y = percent, fill = member_casual)) +
  geom_col(position = "dodge") +
  labs(
    title = "Bike Type Preference by Rider Type (Percentage)",
    x = "Bike Type",
    y = "Percentage By Rider Type",
    fill = "Rider Type"
  ) +
  scale_fill_manual(values=c("casual"="skyblue", member="darkblue")) +
  theme_light() + 
  ylim(0, max(ride_type_percentage$percent) + 5) +
  geom_text(aes(label=sprintf("%.1f%%", percent)),
            position = position_dodge(width = 0.9), vjust = -0.5)
```



